\textit{MSF: Modulated Sub-graph Finder}

### 2019-02-19 MSF_Paper_FA.tex
Updates

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global transcriptional
gene activity changes due to external stimuli or diseases. Differential
gene expression analysis (DGEA) is the first step in the course of data
interpretation, typically producing lists of dozens to thousands of
differentially expressed genes. To further guide the interpretation of
these lists, different pathway analysis approaches have been
developed. These tools typically rely on the classification of genes into
sets of genes, such as pathways, based on the interactions between the genes
and their function in a common biological process. Regardless of technical
differences, these methods do not properly account for cross talk between
different pathways and rely on binary separation into differentially
expressed gene and unaffected genes based on an arbitrarily set p-value
cut-off.

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression patterns
of genes are integrated according to the topology of their interactions and
allow potentially to read the flow of information from the perturbation
source to the effectors. The described software, named \texttt{Modulated
  Sub-graph Finder} (\texttt{MSF}) is freely available at
\url{https://github.com/Modulated-Subgraph-Finder/MSF}.

@@

High throughput sequencing techniques have been widely used to yield
differentially expressed genes (DEG)~\cite{DEG}. To this end, changes in
transcript abundance are measured, e.g.~by next generation sequencing
techniques, and interpreted as an indicator of differential expression of
genes. DEGs can be used to get insights into the mechanism underlying
differences between conditions of samples, such as healthy versus
diseased. Differential gene expression analysis (DGEA) informs about the
magnitude of expression changes between the conditions which are often
expressed as fold change, sign of fold change and the confidence level of
observing an authentic change, often expressed as \textit{p}-value. These
DEGs information is further interpreted to extract meaningful biological
insights. For example, genes that could be involved in the response to a
particular stimuli or maybe the cause of a disease. To this end,
pathway-based analysis has become an important tool to further interpret
the results of a DGEA and to acquire understandings of the perturbations in
a biological system. Biological pathways are sets of genes and their
interactions forming a functional unit. DEGs help to identify pathways or
networks that may be altered during a change of condition providing
important information about diseases and its treatment
process~\cite{Khatri2012}. Pathway-based methods use predefined pathways or
networks such as KEGG~\cite{Kegg} and Reactome~\cite{Reactome}, the
expression measurements of the genes obtained from DGEA in combination with
statistical methods and algorithms to identify specifically modulated
pathways and processes~\cite{Campos}.

Well established resources for pathway annotation are KEGG (Kyoto
Encyclopedia of Genes and Genomes)\cite{Kegg} and
Reactome~\cite{Reactome}. KEGG pathways is a branch of KEGG database that
hosts a collection of manually drawn pathway maps representing the
molecular interaction, reaction and relation networks of cellular
functions. Similarly, Reactome is an open-source, manually curated,
peer-reviewed database for signaling and metabolic molecules with their
interactions formed into different biological pathways for nineteen
species~\cite{Reactome}. Both provide predefined pathways which are sets of
genes and their interactions categorized into functional units. Starting
from a gene interaction network, genes are labeled according their role in
a specific biological process. In this sense a particular gene can be
assigned to different pathways. E.g., the human gene STAT1 is associated
with 24 different pathways in the pathway annotation curated by
KEGG~\footnote{\url{http://www.genome.jp/dbget-bin/www_bget?hsa:6772}} and
in 12 different pathways in the Reactome data
base~\footnote{\url{http://www.reactome.org/content/detail/R-HSA-629622}}.
Although carefully produced, the assignment of genes to those predefined
pathway unites can be considered to be subjective to some degree and
suffers from observational bias~\cite{schnoes2013biases}.

@@

On these grounds we propose a novel approach to make use of the rich gene
and protein interaction annotation resources available to gain additional
functional insights from basic DGEA. To this, we start with the
presupposition that expression of neighboring genes within a functional
pathway are not independent from each other. Rather, they are often
regulating each others expression or are part of the same
regulon~\cite{Michalak}. We understand that the categorization of links
between genes into labeled pathways is often an arbitrary one, given the
extensive cross talk between different pathways. Although these categories
have proven to be useful in many situations, they force a certain
perspective onto the interpretation of novel data. Based on these two
principles, we aim to find sub-graphs of connected genes within cell
signaling network which exhibit as a whole significant differential
expression changes. This approach differs in two main aspects from common
pathway analysis tools. First, it does not aim to identify functional
pathways enriched in differentially expressed genes, but detects sub-graphs
or branches in a network graph (potentially spanning more than one
functionally grouped pathway) which is coherently modulated. Second, it
aims to improve the DGEA on the gene level, by collecting the information
of neighboring genes, which as a whole might exhibit prominent enough
signal to be called, again as a whole, significantly modulated.

@@

All of this can be helpful to understand the cause and effect of a stimulus
and might inform about potential points of intervention. The proposed
algorithm was implemented as a java program, which was named
\texttt{Modulated Sub-graph Finder} (abbreviated
\texttt{MSF}). \texttt{MSF} can help transform the information obtained
from DGEA into comprehensible knowledge of signal transduction of genes and
thereby being a valuable complement to existing pathway based
methods. \texttt{MSF} is freely accessible on github under the terms of the
Creative Commons Attribution 4.0 International License.

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach to find concertedly
modulated sub-graphs in networks of biological interactions.  \texttt{MSF}
does not use predefined gene sets grouped into functional units, but rather
relies purely on the network of interacting genes.  The input network
consists of nodes corresponding to genes and edges representing
interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or
modules, which are as a whole modulated.

\texttt{MSF} uses the individual gene's \textit{p}-values generated from
the DGEA. The \textit{p}-value expresses the probability that the null
hypothesis of unmodified gene expression can't be rejected for a given
statistical model. To find significantly modulated sub-graphs individual
\textit{p}-values of the vicinal genes in the global network are combined
into a single combined \textit{p}-value, using a statistical method for
combining dependent \textit{p}-values described by
Hartung~\cite{Hartung}. Hartung's method uses the inverse of standard
normal distribution function, individual gene \textit{p}-values are first
transformed to their corresponding normal score. Then using these normal
scores, the correlation between genes is calculated, the normal scores and
correlation are applied to the inverse normal function to calculate the
combined \textit{p}-value for all genes examined, namely the examined
sub-graph. The combined \textit{p}-value of a sub-graph will express the
significance of all genes in the sub-graph being modulated
together. Thereby, the information from the different genes are used as,
although not independent, replicated measurement of the behavior of the
whole sub-graph. This potentially increases the power to detect also
significant sub-graphs consisting of genes which are not significant on
there own.
\newline

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four step heuristic as described in the
following. The proceeding identification of modulated sub-graphs from a
network by \texttt{MSF} is presented as a flowchart diagram
(Fig.~\ref{fig:pseudocodemsf}).

\textbf{Initial Modulated sub-graphs}

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed it tries to extend the sub-graph by adding directly
neighboring genes, starting with the most significant one. A single
combined \textit{p}-value is calculated for the extended sub-graph.
If the combined \textit{p}-value is smaller than the \textit{p}-value of
the original sub-graph,
the extended sub-graph is accepted. This step is
iteratively repeated until no further extension is accepted. In this case
the process starts over with all remaining genes not yet in a significantly
modulated sub-graph. This step identifies all simple sub-graphs that
are modulated in the whole network.\newline

\textbf{Extending Modulated sub-graphs}

In the next step, we check if any of the initial modulated sub-graphs 
could further be extended by adding more than one gene at a time. This is done
by testing all possible extension paths up to \emph{N} (default 2) genes at
all nodes in the sub-graph. Again, this step is iteratively repeated until
no further genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

\textbf{Merging Modulated sub-graphs}

After detection and extension of the modulated sub-graphs, they are tested
if combined sub-graphs score better than on their own. The merging of the sub-graphs is done by depth first search traversal from the first sub-graph to the second sub-graph. If the two sub-graphs merge
with the connector of at most \emph{N} genes (default 1 gene) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

\textbf{Finding Sources \& Sinks}

In a last post processing step \texttt{MSF} identifies the trigger points
of the modulated sub-graphs. These trigger genes are the sources of the
sub-graphs with only outgoing edges. These genes can be interpreted as the
possible entry points of perturbation from where the stimulus causes
downstream effects. In the same spirit the most downstream genes of the
modulated sub-graph are identified and defined as sinks. Sinks can be
interpreted as the effectors where the integrated information within the
signal transduction network is set to action. Due to circular loops not all
sub-graphs are guaranteed to have sources or sinks.

@@

\subsection*{Case Study}

To demonstrate its usefulness, \texttt{MSF} is applied to an RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the
Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola. Thereby facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
correctness and usefulness of the approach.

EBOV infection count data was downloaded from GEO (GSE84188). Differential
gene expression analysis was performed on the count data with edgeR package
(version 3.4.2)~\cite{edgeR}. The DEG analysis results generated by edgeR
were used as input for \texttt{MSF}. Directed cell signaling interactions
were filtered from Reactome Functional interactions (FIs) Version
2016~\cite{Cytokegg}, which was used as a second input for \texttt{MSF}.

The EBOV infection experiments describe the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). For the earliest time
point at 6~hpi, three large modulated sub-graphs were identified with 41,
107, and 18 genes were identified, as well as two with less than 4 genes.
The modulated sub-graphs consist predominantly of cytokines, chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23).
IFNB1 and IFNA1 were both identified as two of the possible sources
in one of the sub-graph identified with 18 genes.
Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes (Supplement material 6H).
At 24~hpi eight modulated sub-graphs were
identified with four main sub-graphs consisting
of 27, 101, 134 and 194 genes, others being smaller than 10 genes. IFNB1
and IFNA1 were identified as the two sources out of the total sources. For the last time-point 48~hpi,
six modulated sub-graphs were identified. Three of the sub-graphs were
less than nine genes and main sub-graphs had 122, 191 and 202 genes. IFNB1 was identified as one of the sources in
the most significantly modulated sub-graph (Supplement).

As stated earlier IFN-$\alpha / \beta$ was reported to be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1 as a
source in all three time-points and INFB1 in two of the
time-points. Although IFNA1 and IFNB1 were already among the most
significant gene in the DGEA during the later time points, \texttt{MSF} was
also able to detect them as a source in the very early time-point when the
genes were not significant based on the individual DGEA alone. Identifying
the possible sources will reduce the search space for potential target
genes and can help the biologist as the starting point of clinical testing
for drugs and vaccines against an infection.

Table~\ref{tab:rawVsHartung} compares the results of \texttt{MSF}, namely
the number of detected sub-modules and their total genes numbers, to a
simple analysis of mapping significantly modulated genes from the DGEA to
the network and joining neighbors to modules. The numbers indicate that
\texttt{MSF} detects less but larger sub-modules, applying its statistical
test.  Furthermore, the dependency of the results from the p-value cutoff
choice is demonstrated for the DGEA, which is avoided for \texttt{MSF}
altogether.

@@

In more detail, sub-graph~1 (top) shows how the activation of toll-like
receptor, cytokine, chemokine and jak-stat genes lead via TNF into
apoptosis. The next significant sub-graph (sub-graph~2: middle) reveals
how information from the Extra-cellular matrix (ECM) receptor, which are
reported to interact with Ebola glycoprotein (GP)~\cite{Veljkovic},
chemokines and cytokines, and cytosolic DNA sensing, is integrated into
again modulation of apoptosis pathway. Eventually, sub-graph~3 (bottom)
demonstrate how INFA1 and INFB1 modulate once more, via only a few
intermediate steps, the apoptotic response of the cell.

This show cast example might advertise with how little effort complex data
can be interpreted, help to apprehend the dynamics of the underlying
processes and suggest testable hypothesis and potential points of
intervention.

@@

A potential concern is how noise in the gene expression measurements
affects our analysis. To assess the robustness and stability of our method,
we therefore added Poisson distributed noise
to the read counts of the three time-points data set, used above.
Then DGEA was carried out on the disturbed data
with the same parameters as for the native data using edgeR, followed by
analysis with \texttt{MSF}. This procedure was carried out 100 times.
Every time the genes from the modulated sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the native
data. This was also done for the DEG obtained for each run, using three different cutoffs of
FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
for the time-point 6, 24, and 48~hpi are shown in
Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
considered as rather stringent, which is already reflected by the limited
recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
(24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
pure DGEA can be explained by the fact that disturbed p-values do not
change the results for DGEA as long as the p-value does not rise above the
chosen cutoff value. In contrast, \texttt{MSF} is sensitive to p-value
changes across the whole range of possible values.

\subsection*{Comparison to Reactome pathway analysis tool}

Gene enrichment analysis was performed using Reactome Analyze data
tool~\cite{Reactome}. Reactome's over-representation analysis tests whether
certain Reactome pathways are enriched by the list of genes submitted.Genes from \texttt{MSF} identified
sub-graphs for each time-point were analyzed for gene enrichment analysis.  For
comparison the DEG results from edgeR were filtered using three different cut off of
adjusted p-value 0.01, 0.05 and 0.1. This three subsets of DEG list were used for gene
enrichment analysis. The compression of \texttt{MSF}  and the three subset DEG list is shown in Fig~\ref{fig:msfvsreactome}

The comparison shows most of the pathways known from literature
to be dis-regulated by Ebola infection are enriched in both the enrichment
analysis. Toll-Like receptor signaling pathway when interacts with EBOV
glycoprotein (GP), it triggers the activation of
cytokines~\cite{Olejnik}. Toll-like receptor pathway is expected to be dis-regulated in the early stage of infection,  this pathway was not identified as significantly dis-regulated when p-value
cut off DEG lists were analyzed for enrichment. Ten toll-like receptor cascades were identified as dis-regulated from gene enrichment analysis of \texttt{MSF} identified genes, not a single one of these pathway was shown to be dis-regulated in DEG cut off lists. Since \texttt{MSF}
considers the complete DEG results, even the weak signal at the earliest
time-point was detected; for example Toll-like receptor signaling.

>>>>

High throughput techniques such as RNA-seq or microarray analysis have
proven to be invaluable for the characterization of global transcriptional
gene activity changes due to external stimuli or diseases. Differential
gene expression analysis (DGEA) is the first step in the course of data
interpretation, typically producing lists of dozens to thousands of
differentially expressed genes. To further guide the interpretation of
these lists, different pathway analysis approaches have been
developed. These tools typically rely on the classification of genes into
sets of genes, such as pathways, based on the interactions between the genes
and their function in a common biological process. Regardless of technical
differences, these methods do not properly account for cross talk between
different pathways and rely on binary separation into differentially
expressed gene and unaffected genes based on an arbitrarily set \textit{p}-value
cut-off.

To overcome this limitation, we developed a novel approach to identify
concertedly modulated sub-graphs in the global cell signaling network,
based on the DGEA results of all genes tested. Thereby, expression patterns
of genes are integrated according to the topology of their interactions and
allow potentially to read the flow of information and identify the effectors. The described software, named \texttt{Modulated
  Sub-graph Finder} (\texttt{MSF}) is freely available at
\url{https://github.com/Modulated-Subgraph-Finder/MSF}.

@@

High throughput sequencing techniques have been widely used to yield
differentially expressed genes (DEG)~\cite{DEG}. The changes in
transcript abundance are measured, e.g.~by next generation sequencing
techniques and interpreted as an indicator of differential expression of
genes. DEGs can be used to gain insights into the mechanisms underlying differences between conditions of samples, such as healthy versus infection. Differential gene expression analysis (DGEA) informs about the
magnitude of expression changes, which are often
expressed as log-fold change. The sign of log-fold change and the confidence level of
observing an authentic change, often expressed as \textit{p}-value. The information from DEGs is further interpreted to extract meaningful biological
insights. For example, genes that could be involved in the response to a
particular stimulus or may be the cause of an infection. To this end, pathway-based analysis has become an important tool to further interpret
the results of a DGEA and to acquire understandings of the perturbations in
a biological system. These pathway-based methods use predefined pathways or
networks which are sets of genes with their
interactions forming a functional unit. DEGs help to identify pathways or
networks that may be altered during an infection providing
important information about diseases and its treatment
process~\cite{Khatri2012}. The
expression measurements of the genes obtained from DGEA in combination with
statistical methods and the predefined pathways are used to identify specifically modulated
pathways and processes~\cite{Campos}.

Well established resources for pathway annotation are KEGG (Kyoto
Encyclopedia of Genes and Genomes)~\cite{Kegg} and Reactome~\cite{Reactome}. KEGG pathways is a branch of KEGG database that
hosts a collection of manually drawn pathway maps representing the
molecular interaction, reaction and relation networks of cellular
functions. Similarly, Reactome is an open-source, manually curated,
peer-reviewed database for signaling and metabolic molecules with their
interactions formed into different biological pathways~\cite{Reactome}. Both provide predefined pathways which are sets of
genes and their interactions categorized into functional units.

@@

On these grounds we propose a novel approach to make use of the rich gene
and protein interaction annotation resources available to gain additional
functional insights from basic DGEA. To this, we start with the
presupposition that expression of neighboring genes within a functional
pathway are not independent from each other. Rather, they are often
regulating each others expression or are part of the same
regulon~\cite{Michalak}. We understand that the categorization of links
between genes into labeled pathways is often an arbitrary one, given the
extensive cross talk between different pathways. Although these categories
have proven to be useful in many situations, they force a certain
perspective onto the interpretation of novel data. Based on these two
principles, we aim to find sub-graphs of connected genes within cell
signaling network which exhibit as a whole significant differential
expression changes. This approach differs in two main aspects from common
pathway analysis tools. First, it does not aim to identify functional
pathways enriched in differentially expressed genes, but detects sub-graphs
or branches in a network graph (potentially spanning more than one
functionally grouped pathway) which is coherently modulated. Second, it
aims to improve the DGEA on the gene level, by collecting the information
of neighboring genes, which as a whole might exhibit prominent enough
signal to be called; again as a whole, significantly modulated. All of this can be helpful to understand the cause and effect of a stimulus and might inform about potential points of intervention.

@@

 The proposed algorithm is named
\texttt{Modulated Sub-graph Finder} (abbreviated
\texttt{MSF}). \texttt{MSF} can help transform the information obtained
from DGEA into comprehensible knowledge of signal transduction of genes and
thereby being a valuable complement to existing pathway based
methods.  \texttt{MSF} is freely accessible on github under the terms of the
Creative Commons Attribution 4.0 International License.

\section*{Methods}

\texttt{MSF} was implemented as a java program. It is developed as a novel heuristic approach to find concertedly
modulated sub-graphs in networks of biological interactions.  \texttt{MSF}
does not use predefined gene sets grouped into functional units, but rather
relies purely on the network of interacting genes. The input network
consists of nodes corresponding to genes and edges representing
interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or
modules, which are as a whole modulated.

\texttt{MSF} uses the individual gene's \textit{p}-values generated from
the DGEA. The \textit{p}-value expresses the probability that the null
hypothesis of unmodified gene expression can't be rejected for a given
statistical model. To find significantly modulated sub-graphs individual
\textit{p}-values of the vicinal genes in the global network are combined
into a single combined \textit{p}-value, using a statistical method for
combining dependent \textit{p}-values described by
Hartung~\cite{Hartung}. Hartung's method uses the inverse of standard
normal distribution function. Using the inverse normal cumulative distribution function $\Phi^{-1}$, individual gene \textit{p}-values $T_{i}$ are 
transformed to their corresponding normal score $t_{i}=\Phi^{-1}(1-T_{i})$ that is uniformly distributed on (0,1). Then using these normal
scores, the correlation between genes is calculated $Cov(t_{i},t_{j})=\rho$. The normal scores and
correlation are applied to the weighted inverse normal function to calculate the
combined \textit{p}-value $t(\rho)$ for all genes examined, namely the examined
sub-graph
\newline
 \begin{center}
 	$t(\rho)=\frac{\sum_{i=1}^{n}\lambda_i t_{i} }{\sqrt{(1-\rho) \sum_{i=1}^{n} \lambda^{2}+\rho(\sum_{i=1}^{n} \lambda_i)^{2}}}$
 \end{center}

  Lambda $\lambda$ be the weights for each gene. The combined \textit{p}-value  $t(\rho)$ of a sub-graph will express the
significance of all genes in the sub-graph being modulated
together. Thereby, the information from the different genes are used as,
although not independent, replicated measurement of the behavior of the
whole sub-graph. This potentially increases the power to detect also
significant sub-graphs consisting of genes which are not significant on
there own.
\newline

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a four step heuristic as described in the
following. The proceeding identification of modulated sub-graphs from a
network by \texttt{MSF} is presented as a flowchart diagram
(Fig.~\ref{fig:pseudocodemsf}). \newline

\textbf{Initializing modulated sub-graphs}

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed it tries to extend the sub-graph by adding directly
neighboring genes, starting with the next most significant one. A single
combined \textit{p}-value is calculated for the extended sub-graph.
If the combined \textit{p}-value is smaller than the \textit{p}-value of
the original sub-graph,
the extended sub-graph is accepted. This step is
iteratively repeated until no further extension is accepted. In this case
the process starts over with all remaining genes not yet in the significantly
modulated sub-graph. This step identifies all simple sub-graphs that
are modulated in the whole network.\newline

\textbf{Extending modulated sub-graphs}

In the next step, we check if any of the initial modulated sub-graphs 
could further be extended. This is done when the sub-graph could not be extended any more to the direct neighboring gene due to a bad \textit{p}-value, but the neighboring gene to the direct neighboring gene has a good \textit{p}-value. This was required to connect together good \textit{p}-value gene but incorporating a single bad \textit{p}-value gene. If the combined \textit{p}-value of the initial modulated sub-graph and the extension genes is smaller than the \textit{p}-value of the initial sub-graph the extension is accepted. All possible extension paths up to \emph{3} (default 2) genes at
all nodes in the sub-graph are tested. Again, this step is iteratively repeated until
no further genes are added to the significant differentially expressed
sub-graphs. This step bridges small gaps of genes without a clear
differential signal in the DGEA.\newline

\textbf{Merging modulated sub-graphs}

After detection and extension of the modulated sub-graphs, they are tested
if combining the different sub-graphs score better than on their own. The merging of the sub-graphs is done by depth first search traversal from one sub-graph to the other sub-graph. If the two sub-graphs merge
with the connector of at most \emph{3} genes (default 2 gene) and the
combined \textit{p}-value of the merged sub-graph including the bridging
genes in between is less than the individual \textit{p}-values of the two
sub-graphs, the two sub-graphs are merged together to one big modulated
sub-graph. This step is repeated iteratively until no sub-graphs could be
merged.\newline

\textbf{Finding sources \& sinks}

In a last post processing step \texttt{MSF} identifies the trigger points
of the modulated sub-graphs. These trigger genes are the sources of the
sub-graphs with only outgoing edges. These genes can be interpreted as the
possible entry points of perturbation from where the stimulus causes
downstream effects. Each individual source is given an impact score, depending on the number of down stream genes effected in the corresponding sub-graph, equipping with a way to choose the most significant genes for unknown phenotypes or biological processes. In the same spirit the most downstream genes of the
modulated sub-graph are identified and defined as sinks. Due to loops not all
sub-graphs are guaranteed to have sources or sinks.

\textbf{MSF input}

As input \texttt{MSF} requires the results from
a DGEA in the appropriate file format. So
far the raw output from edgeR~\cite{edgeR} and
DESeq2~\cite{love2014moderated} are supported. Furthermore, a gene network
has to be provided in the format of a directed adjacency list.

\textbf{MSF output}

\texttt{MSF} generates a directed network file as an output, with complete directed interactions of all modulated sub-graphs identified. This file could be imported into Cytoscape~\cite{Cyto} for visualization. Second output file is the source \& sink file. It contains for all the modulated sub-graphs, sources and sinks identified with the source impact score. The third output file, source weightage is provided as a node attribute file. It could be imported into Cytoscape to see the impact score of sources as the size of the node and log-fold change could be used to see how the regulation of different gene hubs in the modulated sub-graphs look.

@@

\subsection*{Case study}

To demonstrate its usefulness, \texttt{MSF} is applied to a RNA-seq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to the Filoviridea family: filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host innate
and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Currently, there are no approved antiviral drugs
for the treatment of Ebola virus infection~\cite{Konde,Rhein}.  The initial
targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. It has been
proposed that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. Ebola infection
data was selected to test the approach because it has been well recognized for the last several decades, and vast literature is available for the pathogenesis of Ebola, thereby, facilitating the verification of the results of
\texttt{MSF} with the vast literature present on Ebola
infection. Especially, the detection of IFN-$\alpha / \beta$ as point of
action for the virus, could be considered as an basic indicator of the
correctness and usefulness of the approach.

EBOV infection count data was downloaded from GEO (GSE84188), it describes the course of infection at three
time-points 6, 24 and 48 hour post infection (hpi). Differential
gene expression analysis was performed on the count data with edgeR package
(version 3.4.2)~\cite{edgeR} using an upper-quartile normalization. The DEG analysis results generated by edgeR
were used as input for \texttt{MSF}. Cell signaling interactions
were filtered from Reactome Functional interactions (FIs) Version
2016~\cite{Cytokegg} for only direct interactions, which was used as a second input for \texttt{MSF}.

For the earliest time point at 6~hpi, three large modulated sub-graphs were identified with 42,
139, and 69 genes.
The modulated sub-graphs consist predominantly of cytokines, chemokines
(CXCL10, CCL8, CXCL9, CXCL11, CXCR4, CCR7, CCL4L1, CCL3L1, CCL4, CCL8, CCL20, CCL3, CCL19) and Interleukin genes (IL6, IL27, IL23).
IFNB1 and IFNA1 were both identified as two of the possible sources
in the most significantly modulated sub-graph identified with 42 genes. The impact score of IFNB1 is 14.5\% and IFNA1 is 8.7\% for the sub-graph they belong to.
Most of the sources identified by \texttt{MSF} were
type~I interferon induced genes (Supplement material 6H).
At 24~hpi seven modulated sub-graphs were
identified with four main sub-graphs consisting
of 61, 222, 130 and 242 genes, others being smaller than 6 genes. Again, IFNB1
and IFNA1 were identified as two sources out of the total sources with 3.9\% and 1.6\% impact score. For the last time-point 48~hpi,
six modulated sub-graphs were identified. Three of the sub-graphs were
less than ten genes and main sub-graphs had 217, 224 and 276 genes. IFNB1 and IFNA1 were identified as sources in
the most significantly modulated sub-graph with an impact score of 2.8\% and 3.7\% (Supplement).

As stated earlier IFN-$\alpha / \beta$ was reported to be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1 and IFNB1 as
sources in all three Ebola infection time-points. Although IFNA1 and IFNB1 were already among the most
significant genes in the DGEA during the later time points, \texttt{MSF} was
also able to detect them as a source in the very early time-point when the
genes were not significant based on the individual DGEA alone. Identifying
the possible sources will reduce the search space for potential target
genes and can help the biologist as the starting point of clinical testing
for drugs and vaccines against an infection.

Table~\ref{tab:rawVsHartung} compares the results of \texttt{MSF}, namely
the number of detected sub-modules and their total genes numbers, to a
simple analysis of mapping significantly modulated genes from the DGEA to
the network and joining neighbors to modules. The numbers indicate that
\texttt{MSF} detects less but larger and meaningful sub-modules, applying its statistical
test. Furthermore, the dependency of the results from the \textit{p}-value cutoff
choice is demonstrated for the DGEA, which is avoided for \texttt{MSF}
altogether. It showcases how applying different cut-offs to the \textit{p}-value of genes from edgeR to the sub-graphs identified by \texttt{MSF} breaks the larger sub-graphs to many smaller unconnected sub-graphs, many of which are single genes.

@@

In more detail, sub-graph~1 (bottom) shows how the activation of toll-like
receptor, cytokine, chemokine activating jak-stat and mapk gene, together with TNF they lead into apoptosis. The next significant sub-graph (sub-graph~2: top right) reveals
how information from the Extra-cellular matrix (ECM) receptor, which are
reported to interact with Ebola glycoprotein (GP)~\cite{Veljkovic},
chemokines and cytokines, and cytosolic DNA sensing, is integrated into
again modulation of apoptosis pathway. Eventually, sub-graph~3 (top left)
demonstrate how INFA1 and INFB1 modulate once more, via only a few
intermediate steps, the apoptotic response of the cell. cAMP signaling genes activates platelet genes.

This display case might advertise with how little effort complex data
can be interpreted, help to apprehend the dynamics of the underlying
processes and suggest testable hypothesis and potential points of
intervention.

@@

A potential concern is how noise in the gene expression measurements
affects our analysis. To assess the robustness and stability of our method,
we therefore added Poisson distributed noise
to the read counts of the three time-points data set, used above.
Then DGEA was carried out on the disturbed data
with the same parameters as for the native data using edgeR, followed by
analysis with \texttt{MSF}. This procedure was carried out 100 times.
Every time the genes from the modulated sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the native
data. This was also done for the DEG obtained for each run, using three different cutoffs of
FDR 0.01, 0.05 and 0.1. The robustness of \texttt{MSF} and the DEG analysis
for the time-point 6, 24, and 48~hpi are shown in
Fig.~\ref{fig:DEGvsMSF}. The procedure how data noise was modeled can be
considered as rather stringent, which is already reflected by the limited
recall rate in the edgeR based DGEA, between 68~\% (6~hpi) and 93~\%
(24~hpi). For \texttt{MSF}-analysis the observed median recall rates lay
between 71~\% (6~hpi) and 84~\% (48~hpi). The better performance of the
pure DGEA can be explained by the fact that disturbed \textit{p}-values do not
change the results for DGEA as long as the \textit{p}-value does not rise above the
chosen cutoff value. In contrast, \texttt{MSF} is sensitive to \textit{p}-value
changes across the whole range of possible values.

\subsection*{Benchmark}

\textbf{jActiveModules}

jActiveModules~\cite{jActiveModules} is a plugin in Cytoscape that searches for molecular interaction network to find expression activated sub-networks. The method used to find the expression activated sub-networks is close to the method used in \texttt{MSF}.

The first time-point of Ebola infection data was analyzed using jActiveModule (Version 3.2.1) to compare the modulated sub-graphs identified by \texttt{MSF} and jActiveModule. The input files were same for both tools, number of modules parameter was set to 5. jActiveModule identified 5 modules, the module with the highest pathScore was selected for comparison with \texttt{MSF} identied modulated sub-graphs. The module consisting of a single graph with 314 genes. While \texttt{MSF} identified three directed modulated sub-graphs with 42, 69 and 139 genes. The overlap of the common genes identified between \texttt{MSF} and jActiveModule is shown in Fig~\ref{fig:MSFvsJActiveModule}. It is difficult to state which tool identifies more accurate set of genes as modulated, since there is no golden standard data example to refer. Although jActiveModule identifies single connected module, \texttt{MSF} provides directionality, with the identification of possible perturbation sources of the sub-graphs.

\textbf{Reactome pathway analysis}

Gene enrichment analysis was performed using Reactome analyze data
tool~\cite{Reactome} (version 67). Reactome's over-representation analysis tool tests whether
certain Reactome pathways are enriched for the list of genes submitted to it. Genes from \texttt{MSF} identified
sub-graphs for each time-point were analyzed for gene enrichment using this tool.  For
comparison the DEG results from edgeR for the three time-points were filtered using the cut-off of
adjusted \textit{p}-value > 0.05. This DEG list was used for gene
enrichment analysis. The compression of \texttt{MSF} identified sub-graphs gene list and the DEG list analysis is shown in Fig~\ref{fig:msfvsreactome}
